---
title: "EPIDP8451_HW6_sp4436"
author: "Sukhman Parhar"
date: "2026-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rpart)
library(caret)
library(rpart.plot)
library(pROC)
library(kernlab)
library(dplyr)
library(e1071)
library(NHANES)
```

## Question 1 Loading in the data

```{r}
# load the data
data(NHANES)

# Inspect structure
str(NHANES)

# look at first few rows
head(NHANES)
```

# Question 2 Clean, restrict, and partition the data

```{r, clean}
# Inspect diabetes
diabetes_struct <- str(NHANES$Diabetes)
diabetes_tab    <- table(NHANES$Diabetes, useNA = "ifany")
diabetes_prop   <- prop.table(diabetes_tab)
diabetes_levels <- levels(NHANES$Diabetes)

diabetes_tab
diabetes_prop
diabetes_levels

# Keep required variables
vars_keep <- c(
  "Age", "Race1", "Education", "HHIncome",
  "Pulse", "Diabetes", "BMI", "PhysActive", "Smoke100"
)

nh <- NHANES %>%
  select(all_of(vars_keep)) %>%
  # (C) Light cleaning
  filter(!is.na(Diabetes)) %>%                               
  mutate(
    Diabetes   = factor(Diabetes, levels = c("No", "Yes")),  
    Race1      = factor(Race1),
    Education  = factor(Education),
    HHIncome   = factor(HHIncome),
    PhysActive = factor(PhysActive),
    Smoke100   = factor(Smoke100)
  ) %>%
  drop_na()                                                 

# Outcome balance
table(nh$Diabetes)
prop.table(table(nh$Diabetes))
```

##
Since we are running logistic regression later on I will be removing weight and height to aviod collinearity. This is really only a problem in LG but I want to mantain the same predictors across models. 

```{r, partition}
set.seed(123)

train_idx <- createDataPartition(y = nh$Diabetes, p = 0.70, list = FALSE)

train_data <- nh %>% slice(train_idx)
test_data  <- nh %>% slice(-train_idx)

# Check balance
prop.table(table(train_data$Diabetes))
prop.table(table(test_data$Diabetes))

# Dimensions
dim(train_data)
dim(test_data)
```

## Question 3: Construct 3 prediciton models 

### Classification Tree

```{r}
set.seed(123)

# 10-fold CV
train.control.class <- trainControl(method="cv", number=10, sampling="down")

p_grid <- expand.grid(cp = seq(0.001, 0.3, by = 0.01))

tree.diab <- train(
  Diabetes ~ .,
  data = train_data,
  method = "rpart",
  metric = "Accuracy",
  trControl = train.control.class,
  tuneGrid = p_grid
)

tree.diab
tree.diab$bestTune
tree.diab$results

tree.diab$finalModel %>%
  rpart.plot()

# variable importance
varImp(tree.diab)

# confusion matrix 
confusionMatrix(tree.diab)
```

### SVC

```{r}
set.seed(123)

train.control.svm <- trainControl(method="cv", number=10, classProbs=TRUE, sampling="down")

svm.diab <- train(
  Diabetes ~ .,
  data = train_data,
  method = "svmLinear",
  trControl = train.control.svm,
  preProcess = c("center", "scale"),
  tuneGrid = expand.grid(C = seq(0.001, 2, length = 30))
)

svm.diab$bestTune
svm.diab

# confusion matrix
confusionMatrix(svm.diab)

plot(svm.diab)
```

### Logistic Regression 

```{r}
set.seed(123)

train.control.lr <- trainControl(method="cv", number=10, sampling="down")

log.diab <- train(
  Diabetes ~ .,
  data = train_data,
  method = "glm",
  family = "binomial",
  trControl = train.control.lr
)

log.diab
```

## Question 4 Select optimal model

```{r}
# Compare models
resamps <- resamples(list(
  Tree = tree.diab,
  SVM  = svm.diab,
  Logistic = log.diab
))

summary(resamps)
bwplot(resamps)
```

##
SVM had the highest mean CV and mean Kappa. It is the most optimal model.

### apply test data
```{r}
# Predictions on test set
svm_pred_test <- predict(svm.diab, test_data)

# Confusion matrix
confusionMatrix(svm_pred_test, test_data$Diabetes, positive = "Yes")

# Predicted probabilities
svm_prob_test <- predict(svm.diab, test_data, type = "prob")

# ROC curve
roc_obj <- roc(response = test_data$Diabetes,
               predictor = svm_prob_test[, "Yes"])

# Plot ROC
plot(roc_obj, col = "black", lwd = 2,
     main = "ROC Curve – SVM (Test Set)")

# AUC
auc(roc_obj)
```

##
Based on cross-validation performance, the Support Vector Machine was selected as the optimal model.When applied to the test data we had a model accuracy of 71.3%. This is lower than the NIR but that is due to class imbalance. We had a high sensitivity of 83.2%, meaning we could correctly identify most individuals with diabetes.The specificity was 69.9%, meaning moderate ability to correctly classify non-diabetic individuals. The area under the ROC curve was 0.82 meaning the model is good at differentiating. 

## Question 5 Ethical Considerations of using race
Many of the ethical concerns around race being used in prediction modeling centers around it being a socially constructed category rather than a biological determinant. This is important bc any observed associations are reflecting underlying structural factors like socioeconomic inequality, environmental exposures, healthcare access, and discrimination. If interpreted uncritically, using race may reinforce biological misconceptions or perpetuate disparities, especially if models inform resource allocation, insurance decisions, or clinical care in ways that disadvantage marginalized groups (such as the discussion we had in class about how a health system that wanted a model to choose were to cut funding). However, excluding race may reduce performance or obscure disparities that require intervention. The major challenge is with how we interpret and implement the results of models that use race. It also should not be a reason that we do not try and get better data on the underlying factors associated with 

