---
title: "EPIDP8451_HW6_sp4436"
author: "Sukhman Parhar"
date: "2026-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rpart)
library(caret)
library(rpart.plot)
library(pROC)
library(kernlab)
library(dplyr)
library(e1071)
library(NHANES)
```

## Question 1 Loading in the data

```{r}
# load the data
data(NHANES)

# Inspect structure
str(NHANES)

# look at first few rows
head(NHANES)
```

# Question 2 Clean, restrict, and partition the data

```{r, clean}
# Inspect diabetes
diabetes_struct <- str(NHANES$Diabetes)
diabetes_tab    <- table(NHANES$Diabetes, useNA = "ifany")
diabetes_prop   <- prop.table(diabetes_tab)
diabetes_levels <- levels(NHANES$Diabetes)

diabetes_tab
diabetes_prop
diabetes_levels

# Keep required variables
vars_keep <- c(
  "Age", "Race1", "Education", "HHIncome",
  "Pulse", "Diabetes", "BMI", "PhysActive", "Smoke100"
)

nh <- NHANES %>%
  select(all_of(vars_keep)) %>%
  # (C) Light cleaning
  filter(!is.na(Diabetes)) %>%                               
  mutate(
    Diabetes   = factor(Diabetes, levels = c("No", "Yes")),  
    Race1      = factor(Race1),
    Education  = factor(Education),
    HHIncome   = factor(HHIncome),
    PhysActive = factor(PhysActive),
    Smoke100   = factor(Smoke100)
  ) %>%
  drop_na()                                                 

# Outcome balance
table(nh$Diabetes)
prop.table(table(nh$Diabetes))
```

##
Since we are running logistic regression later on I will be removing weight and height to aviod collinearity. This is really only a problem in LG but I want to mantain the same predictors across models. 

```{r, partition}
set.seed(123)

train_idx <- createDataPartition(y = nh$Diabetes, p = 0.70, list = FALSE)

train_data <- nh %>% slice(train_idx)
test_data  <- nh %>% slice(-train_idx)

# Check balance
prop.table(table(train_data$Diabetes))
prop.table(table(test_data$Diabetes))

# Dimensions
dim(train_data)
dim(test_data)
```

## Question 3: Construct 3 prediciton models 

### Classification Tree

```{r}
set.seed(123)

# 10-fold CV
train.control.class <- trainControl(method="cv", number=10, sampling="down")

p_grid <- expand.grid(cp = seq(0.001, 0.3, by = 0.01))

tree.diab <- train(
  Diabetes ~ .,
  data = train_data,
  method = "rpart",
  metric = "Accuracy",
  trControl = train.control.class,
  tuneGrid = p_grid
)

tree.diab
tree.diab$bestTune
tree.diab$results

tree.diab$finalModel %>%
  rpart.plot()

# variable importance
varImp(tree.diab)

# confusion matrix 
confusionMatrix(tree.diab)
```

### SVC

```{r}
set.seed(123)

train.control.svm <- trainControl(method="cv", number=10, classProbs=TRUE, sampling="down")

svm.diab <- train(
  Diabetes ~ .,
  data = train_data,
  method = "svmLinear",
  trControl = train.control.svm,
  preProcess = c("center", "scale"),
  tuneGrid = expand.grid(C = seq(0.001, 2, length = 30))
)

svm.diab$bestTune
svm.diab

# confusion matrix
confusionMatrix(svm.diab)

plot(svm.diab)
```

### Logistic Regression 

```{r}
set.seed(123)

train.control.lr <- trainControl(method="cv", number=10, sampling="down")

log.diab <- train(
  Diabetes ~ .,
  data = train_data,
  method = "glm",
  family = "binomial",
  trControl = train.control.lr
)

log.diab
```

## Question 4 Select optimal model

```{r}
# Compare models
resamps <- resamples(list(
  Tree = tree.diab,
  SVM  = svm.diab,
  Logistic = log.diab
))

summary(resamps)
bwplot(resamps)
```

